### 2016-03-07 paper/icfp16-submission/semantics.tex
aasdasd

\paragraph{Holes}
\label{sec:holes}
%
Recall that a key challenge in our setting is to find witnesses
that are meaningful and do not arise from choosing values from
irrelevant types.
%
We solve this problem be equipping our term language with a notion
of a \emph{hole}, written \vhole{\thole}, which represents an \emph{unconstrained}
value that may be replaced with any value of an unspecified type \thole.
%
Dually, holes may also appear in types, written \thole, where they may
be thought of as type variables that we will not generalize over.
%

\paragraph{Substitutions}
%
Our semantics ensure the generality of witnesses by incrementally
\emph{refining} holes, filling in just as much information as is
needed locally to make progress (inspired by the manner in
which SmallCheck uses lazy evaluation~\cite{runciman_smallcheck_2008}).
%
We track how the holes are incrementally filled in, by using
(value) \emph{substitutions} that map holes % (denoted by their index)
to values which may, in turn, contain holes.
%
The substitutions let us ensure that we instantiate each hole
with the same (partially defined) value, regardless of the multiple
contexts in which the hole appears, and hence, ensures we can
report a concrete (and general) witness for any (dynamically)
discovered type errors.

\subsection{Semantics}
\label{sec:semantics}
%

@@

%   % \force{v}{\subst{\tsu}{\thole}}{\vsu}{\tsu} & \mbox{if}\ \thole \in dom(\tsu) \\
%   % \triple{v}{\vsu}{\extendsu{\tsu}{\thole}{\typeof{v}}} & \mbox{otherwise} \\
% \end{cases} \\
\force{\vhole{\thole}}{t}{\vsu}{\tsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \triple{v}{\vsu}{\tsu'}    & \mbox{if } v = \lookupsu{\vsu}{\ehole},
                                         \tsu' = \unify{\{\thole, t, \typeof{v}\}}{\tsu}\\
  \triple{\stuck}{\vsu}{\tsu} & \mbox{if } v = \lookupsu{\vsu}{\ehole} \\
  \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{\{\thole, t\}}{\tsu}, v = \gen{t_1}{\tsu'}\\
\end{cases} \\
% \begin{cases}
%   \triple{\lookupsu{\vsu}{\ehole}}{\vsu}{\tsu} & \mbox{if}\ \ehole \in dom(\vsu), \hastype{\lookupsu{\vsu}{\ehole}}{t} \\
%   \triple{\stuck}{\vsu}{\tsu}                  & \mbox{if}\ \ehole \in dom(\vsu) \\
%   \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu}   & v = \gen{t}\\
% \end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \triple{\vsu(i)}{\vsu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\triple{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{n}{\tint}{\vsu}{\tsu}     & \defeq & \triple{n}{\vsu}{\tsu} \\
\force{b}{\tbool}{\vsu}{\tsu}    & \defeq & \triple{b}{\vsu}{\tsu} \\
\force{\efun{x}{e}}{\tfun}{\vsu}{\tsu} & \defeq & \triple{\efun{x}{e}}{\vsu}{\tsu} \\
\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vleaf{t_1}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu} \\
\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu} \\
\force{v}{t}{\vsu}{\tsu} & \defeq & \triple{\stuck}{\vsu}{\tsu}
\end{array}
$$
\caption{Narrowing values}
\label{fig:narrow}
\end{figure*}
%
While a hole may map to a value that \emph{contains} another hole, \eg a
lambda or a tree, it may not map \emph{directly} to another hole, so in
the first case of \forcesym\ we do not need to \forcesym\ the result of the
substitution, the sub-hole will be narrowed when the context demands it.

\paragraph{Generating Values} The (non-deterministic) procedure $\gen{t}{\tsu}$
takes as input a type $t$ and returns a value of that type. For base
types, the procedure returns an arbitrary value. For functions, it returns
a lambda with a \emph{new} hole denoting the return value, and unconstrained
types (denoted by $\thole$) yield fresh holes constrained to have type \thole
(denoted by $\vhole{\thole}$). \ES{note that $\gen{\ttree{t}}{\tsu}$ will
always place holes in the generated nodes, they will be filled in later,
on demand.}

$$
\begin{array}{lcll}
\gensym       & ::  & (t, \tsu) \rightarrow v \\
\gen{\thole}{\tsu}  & \defeq  & \gen{\subst{\tsu}{\thole}}{\tsu}, & \quad \text{if } \thole \in dom(\tsu) \\
\gen{\tint}{\tsu}   & \defeq  & n, & \quad \text{non-deterministic} \\
\gen{\tbool}{\tsu}  & \defeq  & b, & \quad \text{non-deterministic} \\
\gen{\ttree{t}}{\tsu}  & \defeq  & tr, & \quad \text{non-deterministic} \\
\gen{\tfun}{\tsu}   & \defeq & \efun{x}{\vhole{\thole}}, & \quad \text{\ehole and \thole are fresh} \\
\gen{\thole}{\tsu}  & \defeq & \vhole{\thole}, & \quad \text{\ehole is fresh} \\
\end{array}
$$

@@

\paragraph{Primitive Reductions}
Primitive reduction steps --- addition, if-elimination, function
application, and data construction and case analysis --- use \forcesym
to ensure that values have the appropriate type (and that holes are
instantiated) before continuing the computation. Importantly,
beta-reduction \emph{does not} type-check its argument, it only ensures
that $v_1$ is a function.

\paragraph{Recursion}
Fixed-point operators often cannot be typed in static type systems, but
we are not concerned with \emph{assigning} types to terms, rather with
showing that \emph{no type} can be assigned. We are simply executing the
untyped $\lambda$-calculus, which has no issue handling recursion.

@@

% \begin{proof}%[Proof of \autoref{thm:all-reduce}]
%   Simple induction on the evaluation relation.
% \end{proof}
%
\subsection{Generality}\label{sec:soundness}

A key technical challenge in generating witnesses is
that we have no (static) type information to rely upon. Thus, we
must avoid the trap of generating \emph{spurious} witnesses that
arise from picking irrelevant values, when instead there exist
perfectly good values of a \emph{different} type, under which
the program would not go wrong.

\paragraph{Witness Generality}
We now show that our evaluation relation instantiates holes in a
\emph{general} manner. That is, given a function $f$, if we have
$\steps{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,
then for every possible input type $t$, we can find a value
$v$ of type $t$ such that $\eapp{f}{v}$ goes wrong.

@@

\paragraph{The Type of a a Value} The (dynamic) type of a
value $v$ is defined as a function $\typeof{v}$ defined as:
  \[
  \begin{array}{lcll}
    \typeof{n}   & \defeq & \tint & \\
    \typeof{b}   & \defeq & \tbool & \\
    \typeof{\efun{x}{e}} & \defeq & \tfun \\
    \typeof{\vleaf{t}} & \defeq & \ttree{t} \\
    \typeof{\vnode{t}{v_1}{v_2}{v_3}} & \defeq & \ttree{t} \\
    \typeof{\vhole{\thole}} & \defeq & \thole \\
    % \typeof{\eleaf} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
    % \typeof{\enode{v_1}{v_2}{v_3}} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
    % \typeof{e} & \defeq & \thole, & \quad \text{\thole is fresh} \\
  \end{array}
  \]
%
Note that we approximate the type of values rather than
performing traditional type inference, for example
$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun$,
instead of the usual $\tint \rightarrow \tint$.

\paragraph{Dynamic Type Inference}
We can think of the evaluation of \eapp{f}{\vhole{\thole}} as (dynamically)
inferring a \emph{partial type} --- a type that may contain holes ---
for $f$'s input, \ie a partial instantiation of \thole.
%
We can extract this type from an evaluation trace, by repeatedly
applying the final type substitution to \thole until it contains no
holes in the domain of the substitution. We will call this process of
repeated substition \emph{resolving} a hole, and will use
\resolve{\thole}{\tsu} to denote the the resolution of \thole with
respect to \tsu.
%
Formally, we say that if
$\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\estuck}{\vsu}{\tsu}$,
then the \emph{partial input type} of $f$ upto $\trace$, written
\ptype{\trace}{f}, is
$\resolve{\thole}{\tsu}$.
%$\typeof{\subst{\vsu}{\ehole}}$.
% \ES{should we say $\subst{\tsu}{\thole}$ instead? would need a helper function that does the repeated application of \tsu\ until the result has no $\thole \in dom(\tsu)$}
%
%We will omit the subscript when we wish to refer to the final partial
%type, \ie\ at the step where the expression has been reduced to a value
%(or stuck.)

@@

\paragraph{Type Compatibility}
Two types are compatible if there exists a type subsitution that maps both
types to the same type.
% one of them can be obtained by
% suitably filling the holes of the other (\ie by suitably ``instantiating''
% the type variables appearing in the other).
%
Formally, a type $s$ is
\emph{compatible} with a type $t$, written \tcompat{s}{t}, if
$\exists \tsu.\ \subst{\tsu}{s} = \subst{\tsu}{t}$.

\paragraph{Preservation}
Given these two definitions, we show that each evaluation step
refines the partial input type of $f$, \ie preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$ and
$\trace'$ is a single-step extension of $\trace$, % \defeq \trace, \triple{\estuck'}{\vsu'}{\tsu'}$
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  First note that by Lemma~\ref{lem:force-inst} we can immediately
  discharge the \rulename{E-*-Bad} rules as they cannot change
  \ptype{\trace}{f} at all, and are thus trivially
  compatibility-preserving. For the \rulename{E-*-Good} rules we can
  show that, by virtue of \forcesym succeeding, all must preseve
  compatibility.
  % Note that all rules preserve partial types with the exception of when
  % \forcesym\ is called on a hole, in which case we may instantiate the hole with
  % a concrete value.
  % %
  % But $\typeof{\vhole{\thole}} = \thole$, which is compatible with any type.
\end{proof}

\paragraph{Incompatible Values Are Wrong}
%
\emph{Any} value that is \emph{incompatible} with
the partial input type upto trace $\trace$ will cause $f$ to get stuck
in \emph{at most} $k$ steps, where $k$ is the length of $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  If \steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu} and
     \hastype{v}{t} and
     \tincompat{t}{\ptype{\trace}{f}},
  then
     \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
     in at most $k$ steps, where $k$ is the length of $\trace$.
\end{lem}
\begin{proof}
  By induction on $k$, the length of $\trace$. Let $s_{\trace} = \ptype{\trace}{f}$ and
  $\trace' = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e'}{\vsu'}{\tsu'}$
  such that $\step{e'}{\vsu'}{\tsu'}{e}{\vsu}{\tsu}$.
  Suppose \hastype{v}{\tincompat{t}{s_{\trace}}}, we
  will show that \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
  in at most $k$ steps.

  The base case where $k = 0$ (\ie empty traces) is trivial as no
  evaluation has happened and thus there can be no values that are
  incompatible with \ptype{0}{f}. So we proceed directly to the
  inductive case and split cases on whether $t$ is compatible with
  $s_{\trace'}$.
  %
  \begin{description}
  \item [Case \tincompat{t}{s_{\trace'}}:]
    The inductive hypothesis applies.
  \item [Case $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$:]
    Since $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$ we know
    that $s_{\trace'} \neq s_{\trace}$, and
    by Lemma~\ref{lem:force-inst} we know that we must have
    successfully called \forcesym at step $k$.
    %
    Lemma~\ref{lem:refine-partial} tells us
    $\tcompat{s_{\trace'}}{s_{\trace}}$, which means we must have
    specifically narrowed $s_{\trace'}$ to a type incompatible with $t$.
    %
    A case analysis of the evaluation rules shows that narrowing to
    $t$ instead cannot succeed.
  \end{description}
\end{proof}

>>>>

\paragraph{Holes}
\label{sec:holes}
%
Recall that a key challenge in our setting is to find witnesses
that are meaningful and do not arise from choosing values from
irrelevant types.
%
We solve this problem be equipping our term language with a
notion of a \emph{hole}, written \vhole{\thole}, which represents
an \emph{unconstrained} value $\ehole$ that may be replaced with
\emph{any} value of type \thole.
%
Intuitively, the type holes \thole\ can be viewed as type variables
that we will \emph{not} generalize over.
%
A \emph{normalized} value (resp.\ type) is one that is not a hole,
but which may internally contain holes.
%
For example \RJ{add example here}

\paragraph{Substitutions}
%
Our semantics ensure the generality of witnesses by incrementally
\emph{refining} holes, filling in just as much information as is
needed locally to make progress (inspired by the manner in
which SmallCheck uses lazy evaluation~\cite{runciman_smallcheck_2008}).
%
We track how the holes are incrementally filled in, by using
value (resp.\ type) \emph{substitutions} $\vsu$ (resp. $\tsu$)
that map value (resp.\ type) holes to values (resp.\ types).
%
A \emph{normalized} substitution is one whose co-domain comprises
of normalized values (resp.\ types).
%
In the sequel, we will assume and ensure that all substitutions
are normalized.

\paragraph{Resolving Holes}
We \emph{resolve} a hole with respect to a substitution by
transitively applying the substitution as long as it contains
any holes that are defined in the substitution.
%
We write \resolve{\thole}{\tsu} to denote the the resolution
of \thole\ with respect to \tsu.
%
Note that by definition, \resolve{\thole}{\tsu} does not contain
any holes in the domain of \tsu.

\subsection{Semantics}
\label{sec:semantics}
%
\RJ{Add intuition text -- "dynamic Hindley Milner"}
%
The substitutions let us ensure that we consistently instantiate
each hole with the same (partially defined) value or type, regardless
of the multiple contexts in which the hole appears. This ensures we
can report a concrete (and general) witness for any (dynamically)
discovered type errors.

@@

%   % \force{v}{\subst{\tsu}{\thole}}{\vsu}{\tsu} & \mbox{if}\ \thole \in dom(\tsu) \\
%   % \triple{v}{\vsu}{\extendsu{\tsu}{\thole}{\typeof{v}}} & \mbox{otherwise} \\
% \end{cases} \\
\force{\vhole{\thole}}{t}{\vsu}{\tsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \triple{v}{\vsu}{\tsu'}    & \mbox{if } v = \lookupsu{\vsu}{\ehole},
                                         \tsu' = \unify{\{\thole, t, \typeof{v}\}}{\tsu}\\
  \triple{\stuck}{\vsu}{\tsu} & \mbox{if } v = \lookupsu{\vsu}{\ehole} \\
  \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu'} & \mbox{if}\ \tsu' = \unify{\{\thole, t\}}{\tsu}, v = \gen{t_1}{\tsu'}\\
\end{cases} \\
% \begin{cases}
%   \triple{\lookupsu{\vsu}{\ehole}}{\vsu}{\tsu} & \mbox{if}\ \ehole \in dom(\vsu), \hastype{\lookupsu{\vsu}{\ehole}}{t} \\
%   \triple{\stuck}{\vsu}{\tsu}                  & \mbox{if}\ \ehole \in dom(\vsu) \\
%   \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu}   & v = \gen{t}\\
% \end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \triple{\vsu(i)}{\vsu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\triple{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{n}{\tint}{\vsu}{\tsu}     & \defeq & \triple{n}{\vsu}{\tsu} \\
\force{b}{\tbool}{\vsu}{\tsu}    & \defeq & \triple{b}{\vsu}{\tsu} \\
\force{\efun{x}{e}}{\tfun}{\vsu}{\tsu} & \defeq & \triple{\efun{x}{e}}{\vsu}{\tsu} \\
\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vleaf{t_1}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu} \\
\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq & \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'}, \mbox{if}\ \tsu' = \unify{\{t_1, t_2\}}{\tsu} \\
\force{v}{t}{\vsu}{\tsu} & \defeq & \triple{\stuck}{\vsu}{\tsu}
\end{array}
$$
\caption{Narrowing values}
\label{fig:narrow}
\end{figure*}
%
%While a hole may map to a value that \emph{contains} another hole, \eg a
%lambda or a tree, it may not map \emph{directly} to another hole,
As the substitutions are normalized, in the first case of \forcesym\ we
do not need to \forcesym\ the result of the substitution, the sub-hole
will be narrowed when the context demands it.

\paragraph{Generating Values} The (non-deterministic)
$\gen{t}{\tsu}$ in Figure~\ref{fig:gen}, takes
as input a type $t$ and returns a value of that type.
%
For base types, the procedure returns an arbitrary value of
that type.
%
For functions, it returns a lambda with a \emph{new} hole
denoting the return value, and unconstrained types (denoted
by $\thole$) yield fresh holes constrained to have type
\thole (denoted by $\vhole{\thole}$).
%
Note that for an algebraic datatype, $\gen{\ttree{t}}{\tsu}$
will always place holes in the generated nodes; they will be
lazily filled in later, on demand.

\begin{figure}[ht]
$$
\begin{array}{lcll}
\gensym       & ::  & (t, \tsu) \rightarrow v \\
\gen{\thole}{\tsu}  & \defeq  & \gen{\subst{\tsu}{\thole}}{\tsu}, & \quad \text{if } \thole \in dom(\tsu) \\
\gen{\tint}{\tsu}   & \defeq  & n, & \quad \text{non-deterministic} \\
\gen{\tbool}{\tsu}  & \defeq  & b, & \quad \text{non-deterministic} \\
\gen{\ttree{t}}{\tsu}  & \defeq  & tr, & \quad \text{non-deterministic} \\
\gen{\tfun}{\tsu}   & \defeq & \efun{x}{\vhole{\thole}}, & \quad \text{\ehole and \thole are fresh} \\
\gen{\thole}{\tsu}  & \defeq & \vhole{\thole}, & \quad \text{\ehole is fresh} \\
\end{array}
$$
\caption{Generating values}
\label{fig:gen}
\end{figure}

@@

\paragraph{Primitive Reductions}
%
\RJ{Put high-level intuition about how "dynamic HM" is
formalized in op-sem, to set up next few paragraphs}
%
Primitive reduction steps --- addition, if-elimination,
function application, and data construction and case
analysis --- use \forcesym to ensure that values have
the appropriate type (and that holes are instantiated)
before continuing the computation.
%
Importantly, beta-reduction \emph{does not} type-check its
argument, it only ensures that ``the caller'' $v_1$ is indeed
a function.

\paragraph{Recursion}
\RJ{this para appears out of nowhere --- non-sequitur}
Fixed-point operators often cannot be typed in static type
systems, but we are not concerned with \emph{assigning}
types to terms, rather with showing that \emph{no type}
can be assigned.
%
We are simply executing the untyped $\lambda$-calculus,
which has no issue handling recursion.

@@

% \begin{proof}%[Proof of \autoref{thm:all-reduce}]
%   Simple induction on the evaluation relation.
% \end{proof}

\subsection{Generality}\label{sec:soundness}

A key technical challenge in generating witnesses is
that we have no (static) type information to rely upon.
%
Thus, we must avoid the trap of generating \emph{spurious}
witnesses that arise from picking irrelevant values, when
instead there exist perfectly good values of a \emph{different}
type, under which the program would not have gone wrong.
%
We now show that our evaluation relation instantiates holes
in a \emph{general} manner. That is, given a function $f$,
if we have $\steps{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,
then \emph{for every} input type $t$, we can find a value
$v$ of type $t$ such that $\eapp{f}{v}$ goes wrong.

@@

\paragraph{The Type of a a Value} The \emph{dynamic type}
of a value $v$ is defined as a function $\typeof{v}$ shown
in Figure~\ref{fig:typeof}.
%
The types of primitive values are defined in the natural manner.
%
The types of functions are \emph{approximated}, which is all
that is needed to ensure an application does not get stuck, \eg
$$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun$$
instead of $\tint \rightarrow \tint$.
%
The types of (polymorphic) data types are obtained from the
labels on their values.

\begin{figure}[ht]
\[ \begin{array}{lcll}
    \typeof{n}   & \defeq & \tint & \\
    \typeof{b}   & \defeq & \tbool & \\
    \typeof{\efun{x}{e}} & \defeq & \tfun \\
    \typeof{\vleaf{t}} & \defeq & \ttree{t} \\
    \typeof{\vnode{t}{v_1}{v_2}{v_3}} & \defeq & \ttree{t} \\
    \typeof{\vhole{\thole}} & \defeq & \thole \\
    % \typeof{\eleaf} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
    % \typeof{\enode{v_1}{v_2}{v_3}} & \defeq & \ttree{\thole}, & \quad \text{\thole is fresh} \\
    % \typeof{e} & \defeq & \thole, & \quad \text{\thole is fresh} \\
  \end{array} \]
\caption{The \emph{dynamic type} of a value.}
\label{fig:typeof}
\end{figure}

\paragraph{Dynamic Type Inference}
We can think of the evaluation of \eapp{f}{\vhole{\thole}}
as synthesizing a partial instantiation of \thole, and hence,
\emph{dynamically inferring} a (partial) type for $f$'s input.
%
We can extract this type from an evaluation trace, by
\emph{resolving} the \thole\ with the final type
substitution at the end of the trace.
%
Formally, we say that if
$\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\estuck}{\vsu}{\tsu}$,
then the \emph{partial input type} of $f$ upto $\trace$, written
\ptype{\trace}{f}, is $\resolve{\thole}{\tsu}$.

%repeatedly
%applying the final type substitution to \thole until it contains no
%holes in the domain of the substitution. We will call this process of
%repeated substition \emph{resolving} a hole, and will use
%\resolve{\thole}{\tsu} to denote the the resolution of \thole with
%respect to \tsu.

%$\typeof{\subst{\vsu}{\ehole}}$.
% \ES{should we say $\subst{\tsu}{\thole}$ instead? would need a helper function that does the repeated application of \tsu\ until the result has no $\thole \in dom(\tsu)$}
%
%We will omit the subscript when we wish to refer to the final partial
%type, \ie\ at the step where the expression has been reduced to a value
%(or stuck.)

@@

\paragraph{Compatibility}
%
A \emph{type} $s$ is \emph{compatible} with a type $t$, written \tcompat{s}{t},
if $\exists \tsu.\ \subst{\tsu}{s} = \subst{\tsu}{t}$.
%
That is, two types are compatible if there exists a type substitution
that maps both types to the same type.
%
A \emph{value} $v$ is \emph{compatible} with a type $t$, written \vcompat{v}{t},
if $\tcompat{\typeof{v}}{t}$, that is, if the dynamic type of $v$ is
compatible with $t$.

\paragraph{Preservation}
We prove that each evaluation step \emph{refines} the partial input type
of $f$, \ie\ preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$ and
$\trace'$ is a single-step extension of $\trace$, % \defeq \trace, \triple{\estuck'}{\vsu'}{\tsu'}$
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  First note that by Lemma~\ref{lem:force-inst} we can immediately
  discharge the \rulename{E-*-Bad} rules as they cannot change
  \ptype{\trace}{f} at all, and are thus trivially
  compatibility-preserving. For the \rulename{E-*-Good} rules we can
  show that, by virtue of \forcesym succeeding, all must preseve
  compatibility.
  % Note that all rules preserve partial types with the exception of when
  % \forcesym\ is called on a hole, in which case we may instantiate the hole with
  % a concrete value.
  % %
  % But $\typeof{\vhole{\thole}} = \thole$, which is compatible with any type.
\end{proof}

\paragraph{Incompatible Values Are Wrong}
%
\emph{Any} value that is \emph{incompatible} with
the partial input type upto trace $\trace$ will
cause $f$ to get stuck in \emph{at most} $k$
steps, where $k$ is the length of $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  For all $v$,
  if \steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu} and
     \vincompat{v}{\ptype{\trace}{f}},
  then
     \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
     in at most $k$ steps, where $k$ is the length of $\trace$.
\end{lem}
\begin{proof}
By induction on $k$, the length of $\trace$.
%
Suppose {\vincompat{v}{\ptype{\trace}{f}}}.
%
We show that \steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}
in at most $k$ steps.
%
The base case, $k = 0$ is trivial, $\trace$ is empty
and so $\ptype{\trace}{f}$ is a hole that is compatible
with \emph{every} value $v$.
%
In the inductive case, let
$\trace' = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e'}{\vsu'}{\tsu'}$
be the prefix of $\trace$ of length $k-1$.
%
Furthermore, let $s_{\trace} = \ptype{\trace}{f}$ and $s_{\trace'} = \ptype{\trace'}{f}$.
%
Let us split cases on whether $v$ is compatible with $s_{\trace'}$.
%
\begin{description}
\item [Case \vincompat{v}{s_{\trace'}}:]
  The inductive hypothesis applies.

\item [Case $\vcompat{v}{s_{\trace'}}$ but $\vincompat{v}{s_{\trace}}$:]
  Since $\vcompat{v}{s_{\trace'}}$ but $\vincompat{v}{s_{\trace}}$ we know
  that $s_{\trace'} \neq s_{\trace}$.
  By Lemma~\ref{lem:force-inst} we know that we must have
  invoked \forcesym\ at step $k$.
  %
  The Preservation Lemma~\ref{lem:refine-partial} implies that
  $\tcompat{s_{\trace'}}{s_{\trace}}$, which means we must have
  specifically narrowed $s_{\trace'}$ to a type incompatible with $v$.
  %
  A case analysis of the evaluation rules shows that such an
  invocation of \forcesym\ at step $k$
  cannot succeed, \ie\ yields \stuck.
   \RJ{some intuition needed --- seems like key step.}
\end{description}
\end{proof}
